---
title: "Milestone Report - Word Prediction"
author: "Braian Dias"
date: "October 28, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The main purpose of this report is to present an approach to the problem of predicting the next word given n-first words in a text. This could be used, for instance, to predict the next word in a smart keyboard that we naturally use in our everyday.

To do so, a large corpus of text documents have been used, containing text from newspappers, blogs and Twitter in four different languages: English, German, Russian and Finnish.

The model chosen to accomplish the task was the "n-gram" URL model, which according to Wikipedia, is a type of probabilistic languague model for predicting the next item in such a sequence in the form of a (n-1)-order Markov Model.

In the next sections of the document it will be presented the basic structure of the raw data, how it could be transformed to be suitable for use in a predictive text model and the future plans to address the text prediction problem.

# Loading data

The full dataset is a compressed file of approximatedly 574MB, containing raw text from newspappers, blogs and Twitter in four different languages: English, German, Russian and Finnish. The English file with newspapper text, for instance, is 205mb big uncompressed. Therefore, to be able to run this report in a small fraction of time, a sample of the data will be used.
The function to sample the text takes as input the file name, the number of lines that would be read, and the probability of reading a line (where 1 means all lines would be read, 0.5 means only 50% of the the lines will be read, and so on). 

TODO - profanity filter.

```{r sampling_functions}
library(tm)
library(tokenizers)

en_US.blogs.sample <- readLines("sample/en_US/en_US.blogs.txt")
en_US.news.sample <- readLines("sample/en_US/en_US.news.txt")
en_US.twitter.sample <- readLines("sample/en_US/en_US.twitter.txt")
```

For demonstration purposes, only the English texts will be used as source to the sampling function, reading all the lines with the probability of reading a line of 33%, which means approximadetaly one third of the file will be in the resulting sample file.

```{r sampling}
tokens.blogs <- unlist(tokenize_words(en_US.blogs.sample))
tokens.blogs.df <- data.frame(table(tokens.blogs))
head(tokens.blogs.df[order(-tokens.blogs.df$Freq),])
```

# Exploratory Data Analysis

After loading the raw data, some preprocessing is needed in order to store the text in suitable format to better understand the distribution and relationship between words in the text. 
Text will then be stored in a way to fit a 3-gram model, so it will be a three words pair.

```{r data_transformation}
# TODO
```

## Summary statistics
basic statistics were computed to describe the data.

few lines of the text
Most frequent words
Last frequent words

histograms
## Findings


Data is indeed very noise, containing texts from other languages than the labeled one. There are, in the sampled English texts from newspapers, about N words in other languages.

# Next steps
By having the number of occurences of each 3-gram throughout the English news texts, it's possible to determine the probability of each 3-gram to occur. With this probability, it's possible to predict the next most likely word to occur in a sentence.

Having said that, the next steps would be :
1. Follow the 3-gram approach and create an algorithm to predict the next word.
2. Use the algorithm to create a Shiny app that could be used for everyone to predict the next word given a text.

```{r pressure, echo=FALSE}
plot(pressure)
```

